---
title: "SSI R Training Course"
author: "Jeff Lauder"
date: "Friday, January 02, 2015"
output: ioslides_presentation
---
##SSI R Training Course
January 7th, 2015

This course is a very basic introduction to R and its use in statistical
analysis of SSI data. 


##Introduction
    What is R
        -Extremely powerful Statistical software package, as well as 
        code compiler
        -Can run basic statistics within the base package, as well as
        create your own "apps"
        -Almost limitless as far as statistical power and program 
        complexity
        -Not the easiest to get into right off the bat (no interface)
    What is R Studio
        -"Shell" for R that provides an intuitive user interface.
        -Provides buttons and windows for commonly used operations
##Brief tour of R and R Studio
    -R interface
    -R Studio default interface
    -Customizing R Studio (Tools->Global Options)
    
##Code
    -Everything in R is "coded" or "scripted"
    -R is based in the S computing language, with some basic concepts
    and principles taken from Fortran and Basic, with multiple packages
    that use other languages (Python, SQL, etc.)

##Scripting
    -Much like the old scientist's credo was to "record everything",
    often in notebooks, everything done in R should be written out, into
    a "script" that others, as well as R itself, can follow.
    -Example:
```{r}
#1/2/2015
#Comparison of three sets of random numbers
#Generate 3 random number sets
x1<-rnorm(99,35,3)
x2<-rnorm(99,30,5)
x3<-rnorm(99,28,7)
```
##Scripting
```{r}
#Plot histograms of each distribution
hist(x1)
hist(x2)
hist(x3)

#Step 2:
#ANOVA of three random number sets
```

##Coding basics
    Objects and labeling
    -You can call anything by any name! Don't take labels too literally
    -Be careful, however, of using funtions in names (R is a dumb program
    that needs you to tell it what to do very explicitly)
    -Extremely case sensitive!
    -Annoyingly exact! (As with any coding language:
    careful with commas, etc.)
##Coding basics
```{r}
x<-10
x

my.number<-15
my.number

x<-15
x

The.meaning.of.life<-42
The.meaning.of.life
```
##Coding basics
    Symbology
    -"<-" is same as "="
    -Comments start with "#"
    -Follows basic logical operators:> greater than, etc.
    -Important logic rules: 
    -== "exactly equals"
    -!= "does not equal"
    -&& and & "and"
    -| and || "or"
    Standards/Conventions
    -Always indent major changes in code (e.g. functions, subsets, etc.)
    -Use lots of "comments"
##Coding basics
    Basic functions: Base package
```{r}
#Generate Random number set
x=rnorm(100,50,3)

#Display histogram of random numbers
hist(x)
```
##Coding basics
```{r}
#Calculate basic statistics
mean(x)
sd(x)
median(x)

#Or, use "summary()"
summary(x)
```
##Coding basics
Use new variable names to store results
```{r}
x.summary=summary(x)
x.summary
```
##Coding basics
    Basic functions: getting help
```{r}
?hist
?aov
?friedman.test
```
##Coding basics
    Basic functions: Write your own
    -Calculate square root of randomly generated number between 1 and 100,
    then add the square root back to the original number
    -Note: square root function = sqrt, and random number generation can be
    done with "sample(1:100,1)"
##Coding basics
```{r}
my.number<-sample(1:100,1)
my.number2<-sqrt(my.number)
my.number3<-my.number2+my.number
my.number3
```
##Coding basics
Nesting functions:
```{r}
my.number4<-(sqrt(my.number))+my.number
my.number4
```
##Coding basics
Be careful of what you are putting into your function:
```{r}
my.number5<-(sqrt(sample(1:100,1)))+(sample(1:100,1))
my.number5
```
##Importing data
    Example data: mammal trap data (Thanks Julia!)
    Must reference current data location
    Must be CSV file!
    In R:
```{r}
mammals<-read.csv(file="E:/Documents/SSI/Projects/Mammals/All Data/Sheatsley2012.csv")
mammals
```
##Importing data
    R Studio:
    -In "Environment" tab, click "import dataset". Make sure "header" is
    checked.
    -Can rename file within R studio.
##Data frame versus data view within R
    View() shows whole dataframe separately
    Simply typing the dataset name shows the data within the R console
    Now "view"" your dataset
##Data formats
    "Long"" format versus "short"" format
##Sorting/Exploring Data
    Join/Merge/bind
    Useful for joining two datasets that are already same format
```{r}
mammals2<-read.csv(file="E:/Documents/SSI/Projects/Mammals/All Data/Sheatsley2013.csv")

#mammals.all<-rbind(mammals,mammals2)
#Gets error, extra notes column in mammals2 must be dropped
```
##Sorting continued
    Can drop columns using matrix format by "subsetting":
    Subsetting can be done one of two ways:
    -Directly referencing the matrix format [row,column]
    -Using the "subset" function
    For these types of subsets (removing a column, former is easier)
```{r}
mammals2<-mammals2[,-12]

mammals.all<-rbind(mammals,mammals2)
#View(mammals.all)
```
##More on Subsetting   
    Lets say we want to find only trap dates and locations with a certain 
    species, we can do this with subset(dataframe,function)
```{r}
per.man<-subset(mammals.all,Species.code=="perman")
per.man
```
##More on subsetting
    Now we can use this subset to get basic information like total
    capture counts
    We can do this using the length(column name) function
    When referencing a column, use "$"
```{r}
length(per.man$Trap)
```
    Or attach the dataset, but remember to detach
```{r}
attach(per.man)
length(Trap)
detach(per.man)
```
##Missing values
    What do we do with NA or empty cells?
    R interprets blanks as "0"
    Sometimes can be confusing...
    Example: sometimes may interpret a blank as 0, which will influence
    statistics, whereas "NA" will not
    *May* need to manually set blanks to "NA"
##Missing values
    Not always straightforward
    Have to figure out what R is interpreting them as
    *Don't be afraid to Google your problems!* Tons of info out there!
```{r}
mammals.all[is.na(mammals.all)]<-""
```
##Exploratory analysis
```{r}
#install.packages('reshape2')
library(reshape2)
mam.melt<-melt(mammals.all,id=c("SITE","DATE"),measure.vars="Species.code")
mam.final<-subset(mam.melt,value!="x")
freq.table<-table(mam.final$value)
```

##Exploratory analysis continued
```{r}
mam.final$species<-gsub(" ","",mam.final$value,fixed=TRUE)
mam.final2 <- mam.final[which(mam.final$species!='f'&
                                  mam.final$species!='m'&
                                  mam.final$species!='o'&
                                  mam.final$species!=''), ]
#View(mam.final2)
freq.table<-table(mam.final2$species)
freq.table
```
##Exploratory analysis continued
    Plotting
    Lets use a numeric dataset for this exercise
```{r}
x1<-rnorm(100,50,1)
x2<-rnorm(100,63,7)
x3<-rnorm(100,22,19)
x.data<-data.frame(cbind(x1,x2,x3))
#View(x.data)
```
##Exploratory analysis continued
    Plotting commands are simple
```{r}
plot(x1,x2)
```
##Exploratory analysis continued
    Plotting commands are simple
```{r}
plot(x1,x3)
```
##Exploratory analysis continued
    Plotting commands are simple
```{r}
hist(x1)
```
##Exploratory analysis continued
    Changing figure text
```{r}
hist(x1,xlab="Number",ylab="Frequency",main="My Histogram")
```
##Regression
    All linear models (regression, ANOVA, generalized, etc.) follow
    same basic format in R
    model<-lm(response~predictor,data=dataset) for linear reg.
    or
    model<-aov(response~predictor,data=dateset) for ANOVA models
##Regression
```{r}
model1<-lm(x1~x3,data=x.data)
```
Use "summmary()" for results
##Regression results
```{r}
summary(model1)
```
##Different options for including results on plot
Make sure variable order is correct (predictor on x axis, listed first)
```{r}
plot(x3,x1)
abline(model1)
```

##Different options for including results on plot
Adding regression results to plot
Use simple "text" command, with basic coordinates for location
```{r}
plot(x3,x1,main="Regression of x1 on x3")
abline(model1)
r.2<-summary(model1)$r.squared
p.val<-summary(model1)$coefficients[2,4]
text(52,52, labels = bquote(italic(R)^2))
text(55,52,labels="=")
text(60,52,substr(r.2,1,5))
text(53,51.47,labels=bquote(italic(p)))
text(65,51.5,labels="=")
text(60,51.5,substr(p.val,1,5))
```

##Different options for including results on plot
```{r,echo=FALSE}
plot(x3,x1,main="Regression of x1 on x3")
abline(model1)
r.2<-summary(model1)$r.squared
p.val<-summary(model1)$coefficients[2,4]
text(52,52, labels = bquote(italic(R)^2))
text(55,52,labels="=")
text(60,52,substr(r.2,1,5))
text(53,51.47,labels=bquote(italic(p)))
text(65,51.5,labels="=")
text(60,51.5,substr(p.val,1,5))
```

##Advanced visualization
    GGplot2: histograms, lineplots
    Violin plots
```{r}
#install.packages('ggplot2')
library(ggplot2)
```

##Non-parametric statistics
    #Friedman test, wilcoxon rank sign test


